The Unicode Cookbook \linebreak for Linguists

### 2018-03-12 book/chapters/implementation.tex
Resolves issue #29

\noindent We have created some test data\footnote{\url{https://github.com/unicode-cookbook/recipes/tree/master/Basics/sources}} with the German word \textit{Schächtelchen}, which is the diminuitive form of \textit{Schachtel}, meaning `box', `packet', or `carton' in English.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note the in large text the frequency column may signal errors in the input, such as typos, if they occur with a very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \texttt{Grapheme} column and then add additional columns for translitations or comments.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Chapter~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

<<results = 'asis', echo = FALSE>>=
colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_2
        , caption = 'Profile 2 (splitting by codepoints)'
        , label = 'tab:profile2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

<<>>=
# after NFC normalization unicode codepoints have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives yet another structure of the codepoints
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_3
        , caption = 'Profile 3 (splitting by NFC codepoints)'
        , label = 'tab:profile3'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_4
        , caption = 'Profile 4 (splitting by NFD codepoints)'
        , label = 'tab:profile4'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

@@

Not yet everything is correct in this graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behaviour. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

@@

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

@@

<<tidy=FALSE>>=
# Everything perfekt without regular expressions
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = FALSE
        )$strings

@@

\noindent The JIPA recipes uses excerpts from the Northwind and the Sun passages from the Illustrations of the IPA published in the Journal of the International Phonetic Alphabet. Thus the recipe shows how a user might tokenize IPA proper. The Dogon recipe uses fieldwork data from the Dogon languages of Mali language documentation project.\footnote{\url{http://dogonlanguages.org/}} This recipe illustrates how a user might tokenize fieldwork data from numerous linguists using different transcription practices by defining these practices with an orthography profile to make the output unified and comparable.

>>>>

\noindent We have created some test data\footnote{\url{https://github.com/unicode-cookbook/recipes/tree/master/Basics/sources}} with the German word \textit{Schächtelchen}, which is the diminutive form of \textit{Schachtel}, meaning `box', `packet', or `carton' in English.

@@

\noindent Next, we know a bit about German orthography and which characters combine to form German graphemes. We can use the information from our initial orthography profile to hand-curate a more precise German orthography profile that takes into account capitalization (German orthography obligatorily capitalizes nouns) and grapheme clusters, such as <sch> and <ch>. We can use the initial orthography profile above as a starting point (note the in large text the frequency column may signal errors in the input, such as typos, if they occur with a very low frequency). The initial orthography profile can be edited with a text editor or spreadsheet program. As per the orthography profile specifications (see Chapter \ref{orthography-profiles}), we can adjust rows in the \texttt{Grapheme} column and then add additional columns for transliterations or comments.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(called user-perceived characters, see Chapter~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single code points, others are combinations
of two code points (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode code points,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

<<results = 'asis', echo = FALSE>>=
colnames(profile_2) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_2
        , caption = 'Profile 2 (splitting by code points)'
        , label = 'tab:profile2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by code points (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

<<>>=
# after NFC normalization unicode code points have changed
profile_3 <- write.profile(example, normalize = "NFC", sep = "")
# NFD normalization gives yet another structure of the code points
profile_4 <- write.profile(example, normalize = "NFD", sep = "")
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_3) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_3
        , caption = 'Profile 3 (splitting by NFC code points)'
        , label = 'tab:profile3'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

<<results = 'asis', echo = FALSE>>=
colnames(profile_4) <- c('Grapheme', 'Frequency', 'Codepoint', 'Unicode Name')
print(xtable(profile_4
        , caption = 'Profile 4 (splitting by NFD code points)'
        , label = 'tab:profile4'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'H'
  )
@

@@

Not yet everything is correct in this graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behavior. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

@@

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurrence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

@@

<<tidy=FALSE>>=
# Everything perfect without regular expressions
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = FALSE
        )$strings

@@

\noindent The JIPA recipes uses excerpts from \textit{The North Wind and the Sun} passages from the Illustrations of the IPA published in the Journal of the International Phonetic Alphabet. Thus the recipe shows how a user might tokenize IPA proper. The Dogon recipe uses fieldwork data from the Dogon languages of Mali language documentation project.\footnote{\url{http://dogonlanguages.org/}} This recipe illustrates how a user might tokenize fieldwork data from numerous linguists using different transcription practices by defining these practices with an orthography profile to make the output unified and comparable.

### 2018-03-12 book/chapters/introduction.tex
Resolves issue #29

Writing systems have existed for roughly 6000 years, allowing people to exchange
messages through time and space. Additionally, to quickly bridge large geographic
distances, telegraphic systems of communication (from Greek \emph{τῆλε
γράφειν} `distant writing') have a long and widespread history since ancient
times. The most common telegraphic systems worldwide are so-called whistled
languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
signalling by smoke, fire, flags or even change in water levels through
hydraulic pressure have been used as forms of telegraphy.

>>>>

Writing systems have existed for roughly 6000 years, allowing people to exchange
messages through time and space. Additionally, to quickly bridge large geographic
distances, telegraphic systems of communication (from Greek \emph{τῆλε
γράφειν} `distant writing') have a long and widespread history since ancient
times. The most common telegraphic systems worldwide are so-called whistled
languages \citep{Meyer2015}, but also drumming languages \citep{Meyer2012} and
signaling by smoke, fire, flags or even change in water levels through
hydraulic pressure have been used as forms of telegraphy.

### 2018-03-12 book/chapters/ipa_background.tex
Resolves issue #29

After the Kiel Convention in 1989, the working group assisted
the International Phonetic Association in representing the IPA to the
\textsc{international standards organisation} (ISO) and to the \textsc{text
encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
for phonetic symbols, was used in developing writing system descriptions, which
were at the time being solicited for scripts to be included in the new
multilingual international character encoding standards. The working group for
ISO/IEC 10646 and Unicode were two such initiatives.

In the historical context of the IPA being considered for inclusion in 
ISO/IEC 10646, it is important to realize that there were a variety of 
sources (i.e.\ not just from the Association) which submitted character 
proposals for phonetic alphabets. These proposals, including the one from the 
Association via the Kiel working group, were considered as a whole by 
the ISO working groups that were responsible for incorporating a phonetic 
script into the universal character set (UCS). The ISO working groups that 
were responsible for assigning a phonetic character set then made their 
own submissions as part of a review process by ISO for approval based on 
both informatic and phonetic criteria \citep[86]{EslingGaylord1993}.

>>>>

After the Kiel Convention in 1989, the working group assisted
the International Phonetic Association in representing the IPA to the
\textsc{international organization for standardization} (ISO) and to the \textsc{text
encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
for phonetic symbols, was used in developing writing system descriptions, which
were at the time being solicited for scripts to be included in the new
multilingual international character encoding standards. The working group for
ISO/IEC 10646 and Unicode were two such initiatives.

In the historical context of the IPA being considered for inclusion in 
ISO/IEC 10646, it is important to realize that there were a variety of 
sources (i.e.\ not just from the Association) which submitted character 
proposals for phonetic alphabets. These proposals, including the one from the 
Association via the Kiel working group, were considered as a whole by 
the ISO working groups that were responsible for incorporating a phonetic 
script into the universal character set (UCS). The ISO working groups that 
were responsible for assigning a phonetic character set then made their 
own submissions as part of a review process by ISO for approval based on 
both informatics and phonetic criteria \citep[86]{EslingGaylord1993}.

### 2018-03-12 book/chapters/ipa_meets_unicode.tex
Resolves issue #29

\begin{itemize}
	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item The Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all post-alveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

@@

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
Each of these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@

>>>>

\begin{itemize}
	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item The Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all postalveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

@@

\noindent At the end of this chapter we have added a few longish tables summarizing all
159 different unicode code points that form the basis of strict-IPA encoding
(107 letters, 36 diacritics and 16 remaining symbols). We also make these tables available 
online in CSV format.\footnote{\url{https://github.com/unicode-cookbook/cookbook/tree/master/book/tables}} 
Each of these tables shows a typical glyph, and then lists the Unicode Code point,
Unicode Name and IPA description for each symbol. Further, there is a table with 
the additional options for valid-IPA and a table with the additional options for 
widened-IPA.\@

### 2018-03-12 book/chapters/orthography_profiles.tex
Resolves issue #29

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right}, respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then the must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is probably
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, like for example Unicode codepoints, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of the contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as so in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthograpy profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}

>>>>

\begin{enumerate}
	\def\labelenumi{A\arabic{enumi}.} \setcounter{enumi}{4} 
	\item \textsc{A minimal profile consists of a single column} with a header
       called \texttt{Grapheme}, listing each of the different graphemes in a
       separate line. The name of this column is crucial for automatic 
       processing.
	\item \textsc{Optional columns can be used to specify the left and right
       context of the grapheme}, to be designated with the headers \texttt{Left}
       and \texttt{Right}, respectively. The same grapheme can occur multiple
       times with different contextual specifications, for example to
       distinguish different pronunciations depending on the context. 
	\item \textsc{The columns \texttt{Grapheme}, \texttt{Left} and \texttt{Right}
       can use regular expression metacharacters.} If regular expressions are
       used, then the must be specified in the metadata file as such, 
	   and all literal usage of the special symbols, like full stops <.>
       or dollar signs <\$> (so-called \textsc{metacharacters}) have to be
       explicitly escaped by adding a backslash before them (i.e.~use
       <\textbackslash.> or <\textbackslash\$>). Note that any specification of
       context automatically expects regular expressions, so it is probably
       better to always escape all regular expression metacharacters when used
       literally in the orthography. The following symbols will need to be
       preceded by a backslash: {[} {]} ( ) \{ \} | ~+ * . - ! ? \^{} \$ and the
       backslash \textbackslash~itself. 
	\item \textsc{An optional column can be used to specify classes of graphemes},
       to be identified by the header \texttt{Class}. For example, this column
       can be used to define a class of vowels. Users can simply add ad-hoc
       identifiers in this column to indicate a group of graphemes, which can
       then be used in the description of the graphemes or the context. The
       identifiers should of course be chosen such that they do not conflate
       with any symbols used in the orthography themselves. Note that such
       classes only refer to the graphemes, not to the context. 
	\item \textsc{Columns describing transliterations for each graphemes can be
       added and named at will}. Often more than a single possible
       transliteration will be of interest. Any software application using these
       profiles should prompt the user to name any of these columns to select a
       specific transliteration. 
	\item \textsc{Any other columns can be added freely, but will mostly be ignored
       by any software application using the profiles}. As orthography profiles
       are also intended to be read and interpreted by humans, it is often
       highly useful to add extra information on the graphemes in further
       columns, like for example Unicode code points, Unicode names, frequency of
       occurrence, examples of occurrence, explanation of the contextual
       restrictions, or comments. 
 \end{enumerate}

@@

\begin{enumerate}
	\def\labelenumi{B\arabic{enumi}.} 
	\item \textsc{Each line of a profile will be interpreted as a regular
       expression. }Software applications using profiles can also offer to
       interpret a profile in the literal sense to avoid the necessity for the
       user to escape regular expressions metacharacters in the profile.
       However, this only is possible when no contexts or classes are described,
       so this seems only useful in the most basic orthographies. 
	\item \textsc{The \textsc{class} column will be used to produce explicit
       \textsc{or} chains of regular expressions}, which will then be inserted
       in the \texttt{Grapheme}, \texttt{Left} and \texttt{Right} columns at
       the position indicated by the class-identifiers. For example, a class
       called \texttt{V} as a context specification might be replaced by a regular
       expression like:
       \texttt{(a\textbar{}e\textbar{}i\textbar{}o\textbar{}u\textbar{}ei\textbar{}au)}.
       Only the graphemes themselves are included here, not any contexts
       specified for the elements of the class. Note that in some cases the 
       ordering inside this regular expression might be crucial, therefore it must be 
	   specified as so in the metadata file.
	\item \textsc{The \textsc{left} and \textsc{right} contexts will be included
       into the regular expressions by using lookbehind and lookahead}.
       Basically, the actual regular expression syntax of lookbehind and
       lookahead is simply hidden to the users by allowing them to only specify
       the contexts themselves. Internally, the contexts in the columns
       \texttt{Left} and \texttt{Right} are combined with the column
       \texttt{Grapheme} to form a complex regular expression like:\ 
       \texttt{(?\textless{}=Left)Grapheme(?=Right)}. 
	\item \textsc{The regular expressions will be applied in the order as specified
       in the profile, from top to bottom.} A software implementation can offer
       help in figuring out the optimal ordering of the regular expressions, but
       then it should be made explicit in the orthography profile because regular 
	   expressions are executed in order from top to bottom.
\end{enumerate}
